Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
update results

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.0005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC. In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\

>>>>

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.00005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC. In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\
